Rates of true polar wander in convecting planets

### 2014-05-13 tpw_rate.tex
Add some abstract text

\begin{summary}
Abstract goes where?
\end{summary}

>>>>

\begin{summary}
Mass redistribution in the convecting mantle of a planet can cause perturbations in its moment of inertia tensor. Conservation of angular momentum dictates that these perturbations can change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). Although the existence of TPW on Earth is well-verified, its rate and magnitude over geologic time scales remain controversial. Here we present scaling analyses and novel numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors.

Our numerical model couples the angular and linear momentum equations by implementing a free surface and movable mesh in the CIG-sponsored mantle convection code ASPECT. We allow the mantle to deform self consistently to centrifugal forces and update the rotation axis by solving the angular momentum equation at every time step.  Our simulation and scaling indicate that the important parameters are the Rayleigh number, the rotation rate, and the size of relative density fluctuations (i.e. thermal expansivity times the temperature variations). We identify two timescales, one for the longevity of convective fluctuations, and one for the adjustment time of the hydrostatic bulge. These timescales control the rate and magnitude of TPW, as well as the likelihood of hypothesized "inertial interchange" TPW events.

Finally, we discuss implications for large-scale TPW in Earth's past, which has been suggested to be important for life and climate history.
\end{summary}
